\chapter*{\Large \center Acknowledgements}
\begin{center}
\textit{In the name of Allah, the Beneficent, the Most Merciful. All praises be to Allah who gave me the strength, wisdom, and guidance to complete this thesis.}
\end{center}
This has been a long \textit{six-year} journey, and there are many people I would like to thank. First, I would like to thank my PhD supervisor, Rajeev Raman, for patiently supporting and teaching me over these six years. His mentorship and guidance have allowed me to develop both personally and professionally in ways I never imagined.

A huge thank you to my co-supervisor, Dr. Stanley Fung, for his guidance and invaluable support during this journey. I would also like to thank Thomas Erlebach and Fer-Jan de Vries for their support during my GTA duties and panicky moments. To my fellow PhD researchers in the School of Computing and Mathematical Sciences, thank you for sharing time with me and providing helpful advice during our time together.

I would also like to thank the University of Leicester for the Graduate Teaching Assistantship, which supported my research. Likewise, a huge thank you to the Petroleum Technology Development Fund (PTDF) for the financial support provided to allow me to focus on my research. I am also grateful to the Federal University Birnin Kebbi for the study fellowship approved for me.

Last but not least, I would like to thank my parents, brothers, and sisters for their support and constant prayers. My profound gratitude to my wife for her prayers, support, understanding, and enormous patience with me during this journey.