\chapter*{\Large \center Abstract}

When dealing with big data, where storage and query speed are paramount, traditional data structures are fast but have a huge memory footprint. Succinct structures use space close to the information-theoretic lower bound while remaining competitive in query times compared to traditional ones. Achieving space efficiency while supporting dynamic operations efficiently presents a significant challenge. A common naive approach is to rebuild the static data structure for each update, which becomes very expensive as the data structure grows. 

In this work, several innovative dynamic data structures tailored for integer compression are introduced, addressing both the need for compact representation and the ability to handle updates and queries efficiently. The contributions include extensible arrays designed for building append-only Elias-Fano representations that support incremental data additions without requiring complete reconstruction, shown to be competitive with existing solutions.

A practical implementation of the work of Pătraşcu and Thorup \cite{Patrascu_2006} on constant time predecessor search in small sets is also provided, which was initially seen as theoretical with no practical implementation. It is demonstrated that this implementation is not only simple but also competitive in performance. Leveraging this implementation, a practical implementation of a dynamic searchable prefix-sum data structure based on the works of Dietz \cite{Dietz_1989} and Raman \cite{Raman_2001} is proposed. It is shown that these theoretical data structures can now be practically implemented with SIMD instructions widely available on recent architectures.